---
title: "Mixed Effects Models Case Study"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(Lahman))
```

We will use the Baseball dataset `Teams` available in the `Lahman` package in R, with teams having at least 5 years on record.

```{r}
dat <- Teams %>% 
    count(teamID) %>% 
    right_join(Teams, by = "teamID") %>% 
    filter(n >= 5) %>% 
    select(team = teamID, runs = R, hits = H, walks = BB) %>% 
    mutate(hits = scale(hits),
           walks = scale(walks))
head(dat)

# Hits and walks are subtracted by mean and divided by sd
# This is so the iterative estimating procedure for parameters and random effects converges 
```

We will investigate how hits and walks affects total number of runs (data are scaled to improve convergence of model-fitting). 

Here is a plot on the effect of walks on runs:

```{r, warning=FALSE, fig.width=10}
# Walks vs Runs for each individual team
ggplot(dat, aes(walks, runs)) +
    facet_wrap(~team, ncol=7) +
    geom_point(alpha=0.5) +
    geom_smooth(method="lm", se=FALSE, 
                fullrange=TRUE, size=1) +
    theme_bw()
```

### Two extremes: pooling and separating

Use a linear model to determine the effect of `hits` and `walks` on `runs`, on an average `team`. Best to allow each team to have its own intercept. Note the standard error on `hits` and `walks`.

```{r}
lm(runs ~ walks + hits + team, data = dat) %>% 
    tidy() %>% 
    head()
```

How would you determine the effect of `hits` and `walks` _for each individual team_? Note the standard error on `hits` on a representative team.

```{r}
lm(runs ~ walks*team + hits*team, data = dat) %>% 
    tidy() %>% 
    filter(str_detect(term, "hits"))
```

Which model yields a smaller SE?

"Pooled" model: artificially small, since data are actually correlated.

"Separate" model: artificially large, since we're only using data within a particular group.

### An intermediate: mixed effects models

Fit a mixed effects model with random effects on both `hits` and `walks`. Note the SE's are _in between_ the "pooled" and "separate" linear models.

```{r}
# Note that the std error after fitting LME model is slightly higher than the average model, but dramatically lower than the model that fits every team individually

# Test if estimates are significant using Wald? test? statistic values here are all significant
fit_lme <- lmer(runs ~  walks + hits + (walks + hits | team), data = dat)
tidy(fit_lme)
```

Obtain estimates on individual teams:

```{r}
coef(fit_lme)[[1]]
```

Extract the covariance matrix of the fixed effects:

```{r}
# Tells us the covariance matrix of the sampling distribution of B1 B2 B3 
# Note that the B1 B2 and B3 are estimates, which means they have a sampling distribution
vcov(fit_lme)
```

Extract the covariance matrix of the random effects:

```{r}
# This is different from above and tells us the covariance matrix of the random effects?? 
summary(fit_lme)$varcor[[1]]
```

### Questions

1. Order the models in terms of increasing bias (of regression coefficients)
2. Order the models in terms of increasing variance (of regression coefficients).
3. What is the effect of `hits` on `runs` for team ARI, using the LME?
4. What is the exact distribution of effects (slopes) of `hits` on `runs`?
5. What is the SE of the fixed effect of `hits`?
6. A new team has an average number of hits and walks. Their expected number of runs has what exact distribution?

